Online Balanced Graph Partitioning: Optimizations for Large Clusters

### 2020-05-06 paper/disc20clustering.tex
Rewrite of the crucial part of LB

The constructed ground sets can be perfectly partitioned into the clusters (i.e., nodes of each ground set collocated).
This way, the optimal algorithm moves to a perfect partition at the beginning (where requests incur the cost $0$), thus its cost is bounded.
If the algorithm splits any pair of nodes of a ground set, the adversary issue requests to any pair of split nodes.
These two observations imply that the algorithm must eventually collocate all nodes of a ground set to be competitive.

First, we construct a ground set of size $k-1$ on one of the clusters $S$.
Then, in any configuration, there exists an alone node on $S$.
In such setting, we only issue requests between the current alone node from $S$, and some node that was initially collocated with it.
Repeating such requests, almost all the nodes must visit the cluster $S$.
In comparison, \OPT performs one migration.

\begin{theorem}
	\label{th:lowerbound}
	The competitive ratio of any deterministic online algorithm for \OBRP{} is  $(k\cdot \ell - k - \ell)/2$ for any $k\geq 3$ and for any $\ell \geq 2$.
\end{theorem}

@@

	Then, we create a ground set containing this alone node.
	Precisely,
	we create a ground set $\{x_0, y_0\}$, 
	where $x_0$ is the alone node from $S$ and $y_0$ is an arbitrary node that does not reside in $S$.
	\ALG has them split, and the adversary produces external requests until they are collocated.
	The cluster $S$ already contains a ground set of size $k-1$, hence \ALG must collocate $\{x_0, y_0\}$ in another cluster.
	To preserve a feasible partition of nodes into clusters after collocating $\{x_0, y_0\}$,
	\ALG must replace $x_0$ with another single node.

	We continue adding the node that is currently alone in $S$ to a larger ground set.
	Precisely, for an alone node $x$ from $S$, we merge the ground sets of $x$ and the largest ground set $C$ s.t. $C \neq \{x_0,y_0\}$ and $I(C) = I(x)$.
	This way, $x$ is evicted similarly to $x_0$, and we continue creating such ground sets (we describe the termination condition shortly).
	Note that subsequent ground sets are different than the first ground set $\{x_0, y_0\}$: now the nodes originate in the same cluster, whereas nodes $x_0$ and $y_0$ were initially in different clusters.

	We claim that expanding ground sets in this way can continue as long as at least $\ell$ alone node exists.
	Fix the moment just before the creation of a ground set with the node $x_i$.\maciek{no $x_i$ def any more}
	We show that if at least $\ell$ alone nodes exists (including $x_i, y_i$), then there exists a perfect partition of ground sets.

	To show this claim, we consider a configuration, where the largest ground set of each cluster originates from it (excluding the ground set $\{x_0, y_0\}$).
	In this configuration, each cluster contains:  at most one ground set larger than $1$, possibly some alone nodes, and possibly the ground set $\{x_0, y_0\}$.
	If $I(x_i)$ contains an alone node, then we simply obtain a perfect partition by swapping it with $x_i$.
	Otherwise, $\{x_0, y_0\}$ is present on $I(x_i)$ (a single ground set cannot occupy the whole cluster $I(x_i)$, as without $x_i$ its size is at most $k-1$).
	In this case, we first swap $\{x_0, y_0\}$ with two alone nodes that occupy the same cluster.
	To see that such two nodes must exist, observe that we have at most $\ell-1$ clusters with alone nodes, and by the assumption we have $\ell$ alone nodes.
	After this swap, we migrate $x_i$ to $I(x_i)$, evicting a lone node to $S$, and we conclude that the perfect partition exists.

@@

\section{Distributed implementation.}
The algorithm PPL from Section~\ref{sec:ppl} can be distributed
similarly to the approach in~\cite{sigmetrics19_partitioning}.
The algorithm for $k=2$ from Section~\ref{sec:k2} performs only local communication for each request: counters are kept on the clusters and updated locally, and each migration concerns only two clusters.

The algorithm for $k=3$ from Section~\ref{sec:k3} can be efficiently distributed.
There are two non-local operations:
\begin{enumerate}
	\item Broadcasting the end of the phase.\label{it:broadcast}
	\item Component merge.
	\label{it:merge}
\end{enumerate}

>>>>

The constructed ground sets can be perfectly partitioned into the clusters (i.e., with nodes of each ground set collocated).
This way, the optimal algorithm moves to a perfect partition at the beginning (where requests incur the cost~$0$), thus its cost is bounded.
If the algorithm splits any pair of nodes of a ground set, the adversary issues requests to any pair of split nodes.
These two observations imply that the algorithm must eventually collocate all nodes of a~ground set to be competitive.

First, we construct a ground set of size $k-1$ on an arbitrary cluster $S$.
In any configuration, there exists an alone node on $S$.
We issue requests between the current alone node from $S$, and some node that was initially collocated with it.
Repeating such requests, almost all the nodes must visit the cluster $S$.
In comparison, \OPT performs one migration.

\begin{theorem}
	\label{th:lowerbound}
	The competitive ratio of any deterministic online algorithm for \OBRP is at least $(k\cdot \ell - k - \ell)/2$ for any $k\geq 3$ and $\ell \geq 2$.
\end{theorem}

@@

	Then, we add the alone node from $S$ to a larger ground set to force its eviction.
	Precisely,
	we create a ground set $\{x_0, y_0\}$, 
	where $x_0$ is the alone node from $S$ and $y_0$ is an arbitrary node that does not reside in $S$.
	\ALG has them split, and the adversary produces external requests until they are collocated.
	The cluster $S$ already contains a ground set of size $k-1$, hence \ALG must collocate $\{x_0, y_0\}$ in another cluster.
	To preserve a feasible partition of nodes into clusters after collocating $\{x_0, y_0\}$,
	\ALG must replace $x_0$ with another single node.

	We continue adding the node that is currently alone in $S$ to a larger ground set.
	Precisely, for an alone node $x$ from~$S$, we merge the ground sets of $x$ and the largest ground set $C$ s.t. $C \neq \{x_0,y_0\}$ and $I(C) = I(x)$.
	This way, $x$ is evicted similarly to $x_0$, and we continue creating such ground sets.
	We describe the termination condition shortly.
	(Note that these ground sets contain nodes that originate from the same cluster, whereas the nodes $x_0$ and $y_0$ originate from different clusters.)

	We claim that expanding ground sets in this way can continue as long as at least $\ell$ alone node exists.
	Fix the moment just before the creation of a ground set with the alone node $z$ that currently resides in $S$.
	We show that if at least $\ell$ alone nodes exists (including $z$), then there exists a perfect partition of ground sets after expanding a ground set with $z$.

	To show this claim, we consider a configuration, where the largest ground set of each cluster originates from it (excluding the ground set $\{x_0, y_0\}$).
	In this configuration, each cluster contains:  at most one ground set larger than $1$, possibly some alone nodes, and possibly the ground set $\{x_0, y_0\}$.
	If $I(z)$ contains an alone node, then we simply obtain a~perfect partition by swapping it with $z$.
	Otherwise, if no alone nodes exist on $I(z)$, then its largest ground set is at most $k-1$, because $z$ is on $S$.
	Hence, $\{x_0, y_0\}$ is present on $I(z)$, and we must evict it to make space for $z$.
	To find the cluster for $\{x_0, y_0\}$, observe that we have at most $\ell-1$ clusters with alone nodes, and we have at least $\ell$ alone nodes --- hence a cluster with two alone nodes exists.
	To find a feasible partition, we swap $\{x_0, y_0\}$ with these two collocated alone nodes, and then we swap an alone node from $I(z)$ with $z$.

@@

\section{Distributed implementation.}
The algorithm PPL from Section~\ref{sec:ppl} can be distributed
similarly to the approach in~\cite{sigmetrics19_partitioning}.
The algorithm for $k=2$ from Section~\ref{sec:k2} performs only local communication for each request: counters are kept on the clusters and updated locally, and each migration concerns only two clusters.
\medskip

In the rest of this section, we propose the distributed implementation of the algorithm for $k=3$ from Section~\ref{sec:k3}.
The algorithm performs two non-local operations:
\begin{enumerate}
	\item Broadcasting the end of the phase.\label{it:broadcast}
	\item Component merge.
	\label{it:merge}
\end{enumerate}
